---
title: "*Z*-Scores"
subtitle: "PSYC 2020-A01 / PSYC 6022-A01 | Lab 4"
author: "Shuheng Jiang"
format:
  live-revealjs:
    slide-number: true
    show-slide-number: all
    css: styles.css
    width: 1200
engine: knitr
webr:
  packages:
    - ggplot2
execute:
  echo: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

## Outline

* *Z*-Scores
* *Z*-Scores in R
* Generating Data
* Probabilities of Normal Values

Learning objectives:
**R:** Generating data, normal distribution functions

**Statistics:** *Z*-scores

## *Z*-Scores

To "*Z*-score" or "standardize" your data
<div class="subpoint">○ Transform a variable such that the mean is zero and the standard deviation is one </div>
<div class="subpoint">○ Matches the mean and SD of a "standard" normal distribution, $N(0, 1)$ </div>

```{r, echo=F}
#| fig-width: 4
#| fig-height: 3
#| layout-ncol: 3
library(ggplot2)

ggplot(data = iris, aes(x = Sepal.Width)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10) +
  stat_function(inherit.aes = FALSE, fun = dnorm,
                n = 101, args = list(mean = 0, sd = 1),
                color = "lightblue", linewidth = 2,
                xlim = c(-3, 3)) +
  coord_cartesian(xlim = c(0, 6), ylim = c(0, 1)) +
  theme_classic()

ggplot(data = iris, aes(x = Sepal.Width - mean(Sepal.Width))) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10) +
  stat_function(inherit.aes = FALSE, fun = dnorm,
                n = 101, args = list(mean = 0, sd = 1),
                color = "lightblue", linewidth = 2,
                xlim = c(-3, 3)) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(0, 1)) +
  theme_classic()

ggplot(data = iris, aes(x = (Sepal.Width - mean(Sepal.Width)) / sd(Sepal.Width))) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10) +
  stat_function(inherit.aes = FALSE, fun = dnorm,
                n = 101, args = list(mean = 0, sd = 1),
                color = "lightblue", linewidth = 2,
                xlim = c(-3, 3)) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(0, 1)) +
  theme_classic()
```

## *Z*-Scores

Why standardize?

* Puts variables on the same (interpretable) scale

* Helps manage very large or very small numbers

* Can compare across distributions

* Gives information about location relative to the rest of the distribution

## *Z*-Scores in R

Need to:

1. Take out the mean

2. Scale by the standard deviation

`data$variable_z <- (data$variable - mean(data$variable)) / sd(data$variable)`


```{webr}
# let's z-score Petal.Length

```

::: fragment
Can check to make sure things look right
:::

## Generating Normal Data {style="font-size:85%"}

Sometimes, we want to simulate data

<div class="subpoint">○ Demonstration </div>
<div class="subpoint">○ Comparison </div>
<div class="subpoint">○ Sanity check </div>

To generate data drawn from a normal distribution, we can use `rnorm(n, mean, sd)`
<div class="subpoint">○ `n` = number of draws </div>
<div class="subpoint">○ `mean` = desired mean of distribution </div>
<div class="subpoint">○ `sd` = desired SD of distribution </div>

```{r, echo=F}
#| fig-width: 2.5
#| fig-height: 2.5
#| layout-ncol: 3

ggplot(data = data.frame(x = rnorm(10, 0, 1)), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10) +
  stat_function(inherit.aes = FALSE, fun = dnorm,
                n = 101, args = list(mean = 0, sd = 1),
                color = "lightblue", linewidth = 2,
                xlim = c(-3, 3)) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(0, 1)) +
  labs(title = "n = 10") +
  theme_classic()

ggplot(data = data.frame(x = rnorm(50, 0, 1)), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10) +
  stat_function(inherit.aes = FALSE, fun = dnorm,
                n = 101, args = list(mean = 0, sd = 1),
                color = "lightblue", linewidth = 2,
                xlim = c(-3, 3)) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(0, 1)) +
  labs(title = "n = 50") +
  theme_classic()

ggplot(data = data.frame(x = rnorm(500, 0, 1)), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 10) +
  stat_function(inherit.aes = FALSE, fun = dnorm,
                n = 101, args = list(mean = 0, sd = 1),
                color = "lightblue", linewidth = 2,
                xlim = c(-3, 3)) +
  coord_cartesian(xlim = c(-3, 3), ylim = c(0, 1)) +
  labs(title = "n = 500") +
  theme_classic()

```

## Generating Normal Data

```{webr}
# let's generate some data and then plot it

```

## Probabilities of Normal Values

Within a normal distribution, we can check the probability of a given value and the value at a given probability

:::: {.columns}
::: {.column}
#### Probability of a Value
`pnorm(q)`
<div class="subpoint">○ `q` = value (or vector of values) </div>
```{r}
pnorm(0)
pnorm(1)
pnorm(c(-3, 3))
```
:::

::: {.column}
#### Value at a Probability
`qnorm(p)`
<div class="subpoint">○ `p` = probability (or vector of probabilities) </div>
```{r}
qnorm(.5)
qnorm(.9)
qnorm(c(.025, .975))
```
:::
::::

## Visualize: `ggplot2`

So many numbers! Can we please visualize our data?

::: {style="font-size:75%;"}
> "The simple graph has brought more information to the data analyst's mind than any other device." — John Tukey
:::

Yes! With the `ggplot2` package

A bit of a different plotting system than the base R plots we've used

But much more expansive and flexible (and much nicer looking, if you ask me!)

## Visualize: Setup {style="font-size:85%"}

Run this cell first — everything on the next slides builds on it

<div class="subpoint">○ Loads `ggplot2` </div>
<div class="subpoint">○ Adds a grouping variable to `iris` with `cut()`, just like we did in Lab 3 </div>

```{webr}
#| autorun: true
library(ggplot2)

iris$Size <- cut(iris$Sepal.Width,
                 breaks = c(0, 3, 10),
                 labels = c("Narrow sepal", "Wide sepal"))

head(iris)
```

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns}
::: {.column width=66%}
```{r, echo=F}
#| fig-width: 6
#| fig-height: 4
library(ggplot2)
iris$Size <- cut(iris$Sepal.Width, breaks = c(0, 3, 10),
                 labels = c("Narrow sepal", "Wide sepal"))

ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species, fill = Species)) +
  geom_point(size = 2.4) +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0, 8), ylim = c(0, 8)) +
  labs(x = "Sepal Length (cm)", y = "Petal Length (cm)") +
  facet_wrap(~ Size) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom",
        aspect.ratio = 1)
```
:::

::: {.column width=33%}
Let's begin at the end:

This is where we're headed. We'll build it up one layer at a time.
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris)
```
:::

::: {.column width=33%}
When we just call the `ggplot()` function with our data as the first argument, we get this blank area

`ggplot()` works by adding layers and layers onto a plot object

> This is not a very exciting plot, but you can think of it like an empty canvas you'll paint the remaining layers of your plot onto. — R4DS
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris,
       mapping = aes(x = Sepal.Length, y = Petal.Length))
```
:::

::: {.column width=33%}
Next, we want to tell `ggplot()` what information we want in our plot and where we want it

We do this with the `aes()` ("aesthetics") function in the `mapping =` argument

How variables are *mapped* onto specific parts of the plot

To investigate the relationship between sepal length and petal length, we can put `Sepal.Length` on the x-axis and `Petal.Length` on the y-axis

What do we get?
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris,
       mapping = aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point()
```
:::

::: {.column width=33%}
Still no data, though

Need to define some **geom**, a geometric display of our data

Will usually start with `geom_`

<div style="font-size:70%">
|   |   |
|---|---|
| `geom_line()` | line graph |
| `geom_bar()` | bar graph |
| `geom_boxplot()` | boxplot |
| `geom_point()` | scatterplot |
</div>

Can start addressing questions here!
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris,
       mapping = aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 8), ylim = c(0, 8))
```
:::

::: {.column width=33%}
These variables are conceptually similar (both lengths in cm) and should probably have the same scale

Automatically generated axes are a good place to start

But good practice to consider them each time

Try changing the limits and re-running!
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris,
       mapping = aes(x = Sepal.Length, y = Petal.Length,
                     color = Species)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 8), ylim = c(0, 8))
```
:::

::: {.column width=33%}
Adding aesthetics and layers

Which of these flowers belong to which species?

Start: do we need to modify the aesthetic or the geom?

What can we tell from this?
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris,
       mapping = aes(x = Sepal.Length, y = Petal.Length,
                     color = Species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0, 8), ylim = c(0, 8))
```
:::

::: {.column width=33%}
Adding aesthetics and layers

Let's say we want to add a line to better represent the overall trend

Start: do we need to modify the aesthetic or the geom?

What can we tell from this?
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris,
       mapping = aes(x = Sepal.Length, y = Petal.Length,
                     color = Species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0, 8), ylim = c(0, 8)) +
  facet_wrap(~ Size)
```
:::

::: {.column width=33%}
Let's say we want to look at this separately for narrow-sepal vs wide-sepal flowers

`facet_wrap()` splits the plot into panels by a grouping variable

What can we tell from this?
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris,
       mapping = aes(x = Sepal.Length, y = Petal.Length,
                     color = Species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0, 8), ylim = c(0, 8)) +
  facet_wrap(~ Size) +
  theme_classic()
```
:::

::: {.column width=33%}
Let's say we want this plot to look nicer!!

Built in themes

<div class="subpoint">○ Try `theme_minimal()` or `theme_bw()` instead </div>
:::
::::

## Visualize: `ggplot2`

Our question: what is the relationship between sepal length and petal length?

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length,
                 color = Species, fill = Species)) +
  geom_point(size = 2.4) +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0, 8), ylim = c(0, 8)) +
  labs(x = "Sepal Length (cm)", y = "Petal Length (cm)") +
  facet_wrap(~ Size) +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom",
        aspect.ratio = 1)
```
:::

::: {.column width=33%}
Let's say we want this plot to look nicer!!

Additional theming with `theme()`, `labs()`, and various arguments

And we're back to the plot we started with!
:::
::::

## Visualize: Boxplots

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
ggplot(data = iris,
       mapping = aes(x = Species, y = Sepal.Length,
                     fill = Species)) +
  geom_boxplot() +
  labs(x = "Species", y = "Sepal Length (cm)") +
  theme_classic()
```
:::

::: {.column width=33%}
`geom_boxplot()` shows a distribution for each group

<div class="subpoint">○ `x` = the grouping variable (categorical) </div>
<div class="subpoint">○ `y` = the variable you're describing (numeric) </div>

Same anatomy as before: median, Q1, Q3, whiskers, outliers

Things to try:
<div class="subpoint">○ Swap `x` and `y` to flip it sideways </div>
<div class="subpoint">○ Add `geom_jitter(width = .1)` to show the raw data points on top </div>
:::
::::

## Visualize: Barplots

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
# geom_bar() counts rows for you
ggplot(data = iris,
       mapping = aes(x = Size, fill = Species)) +
  geom_bar() +
  labs(x = "Sepal width group", y = "Count") +
  theme_classic()
```
:::

::: {.column width=33%}
Two different bar geoms, and the difference matters:

`geom_bar()`
<div class="subpoint">○ Counts the rows in each category for you </div>
<div class="subpoint">○ Only needs an `x` </div>

`geom_col()`
<div class="subpoint">○ Uses a value you already computed </div>
<div class="subpoint">○ Needs both `x` and `y` </div>

Things to try:
<div class="subpoint">○ Add `position = "dodge"` inside `geom_bar()` to put bars side by side </div>
:::
::::

## Visualize: Barplots with `geom_col()`

:::: {.columns style="font-size:60%"}
::: {.column width=66%}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
# first compute the means, then plot them
mean_lengths <- aggregate(Sepal.Length ~ Species,
                          data = iris,
                          FUN = mean)
mean_lengths

ggplot(data = mean_lengths,
       mapping = aes(x = Species, y = Sepal.Length,
                     fill = Species)) +
  geom_col() +
  labs(x = "Species", y = "Mean Sepal Length (cm)") +
  theme_classic()
```
:::

::: {.column width=33%}
Here we use `aggregate()` from Lab 3 to get one number per group, then `geom_col()` draws those numbers directly

A note on bar plots of means:
<div class="subpoint">○ They hide the spread of the data </div>
<div class="subpoint">○ A boxplot usually tells you more </div>
:::
::::

## Lab 4 Assignment!

<!-- Add assignment reminders here. -->
- Please submit both the knitted file and the rmd. on Canvas!
