---
title: "Sampling Distributions & the Central Limit Theorem"
subtitle: "PSYC 2020-A01 / PSYC 6022-A01 | Lab 6"
author: "Shuheng Jiang"
format:
  live-revealjs:
    slide-number: true
    show-slide-number: all
    css: styles.css
    width: 1200
engine: knitr
execute:
  echo: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{webr}
#| edit: false
#| output: false
# hidden setup: runs when the page loads so every live cell below has the U-shaped data
Ushape <- c(rep(5, 250), rep(4.5, 200), rep(4, 100), rep(3.5, 50), rep(3, 25),
            rep(2.5, 50), rep(2, 100), rep(1.5, 200), rep(1, 250))
```

```{r}
#| include: false
Ushape <- c(rep(5, 250), rep(4.5, 200), rep(4, 100), rep(3.5, 50), rep(3, 25),
            rep(2.5, 50), rep(2, 100), rep(1.5, 200), rep(1, 250))
```

## Outline

-   Population Parameters & Estimation
-   Sampling Distributions
-   Central Limit Theorem
-   Sampling Distributions in R

Learning objectives: **R:** `sample()`, `set.seed()`, `for` loops, `rnorm()` / `runif()` / `rexp()`

**Statistics:** Population vs. sample vs. sampling distributions, Central Limit Theorem, standard error

# Population Parameters {.center}

## What does a researcher want to know? {style="font-size:85%"}

| Data (what we have) | Research question is about… (what we want to know) |
|----|----|
| Randomly selected 1000 people's IQ in Georgia | IQ of people living in GA |
| 30 students' height in the Psych department | Height of students in Psych |
| Randomly selected 100 GT students' sleeping hours | Sleeping time of students at GT |
| Product ratings from 300 people | Product satisfaction for customers |

::: fragment
We are interested in *population* properties (i.e., population **parameters**)

We infer information about the population from the data we have
:::

## Population Parameters

Population parameters are…

::: subpoint
○ Assumed to be constant (fixed)
:::

::: subpoint
○ Unknown and unknowable
:::

::: subpoint
○ Estimated using statistics
:::

::: {style="font-size:80%"}
ⓘ *μ* (mu; mean) and *σ* (sigma; standard deviation) are parameters of the normal distribution; different distributions have different parameters
:::

. . .

Statistics is the science of testing hypotheses about values of these (and other) parameters

::: subpoint
○ Point estimation
:::

::: subpoint
○ Interval estimation
:::

## Two Types of Estimation {style="font-size:85%"}

::::::::: columns
::::: column
**Point estimation**

A single value (point) estimate of the parameter's true value

::: subpoint
○ e.g., $\hat{\mu} = 52$
:::

::: subpoint
○ $\hat{\mu}$ and $M$ both refer to the sample estimate of the mean
:::
:::::

::::: {.column .fragment}
**Interval estimation**

A range of values that may contain the population parameter

::: subpoint
○ e.g., $\hat{\mu} = (100, 150)$
:::

::: subpoint
○ Usually reported with a confidence level (e.g., 95%) to indicate the method's reliability
:::
:::::
:::::::::

::: {.fragment style="font-size:80%"}
If we drew 100 random samples and computed a 95% confidence interval (CI) for each one, we would expect about 95 of those 100 CIs to include the true population mean
:::

# Sampling Distributions {.center}

## Let's Get a Sample

::::: columns
::: {.column width="55%"}
A *sample* consists of multiple observations drawn from the population

$$X_1, X_2, X_3, \dots, X_{n}$$

We summarize the sample with its **sample mean**

$$\bar{X} = \frac{\sum x}{n}$$
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
par(mar = c(0, 0, 0, 0))
plot(NA, xlim = c(0, 10), ylim = c(0, 10), axes = FALSE, xlab = "", ylab = "", asp = 1)
symbols(2.5, 5, circles = 2.3, inches = FALSE, add = TRUE, bg = "#7fd6a4", fg = "#4a9c6e")
text(2.5, 5, "Population", cex = 1.4)
ys <- seq(9, 1, length.out = 6)
arrows(4.8, 5, 7.3, ys, col = "red", lwd = 2, length = 0.1)
symbols(rep(7.9, 6), ys, circles = rep(0.55, 6), inches = FALSE, add = TRUE, bg = "#7fd6a4", fg = "#4a9c6e")
text(7.9, ys, c(expression(X[1]), expression(X[2]), expression(X[3]),
                expression(X[4]), expression(X[5]), expression(X[n])), cex = 1.1)
```
:::
:::::

## Sampling Distribution

:::::::: columns
:::::: {.column width="55%"}
Multiple means from multiple samples

$$\bar{X}_1, \bar{X}_2, \bar{X}_3, \dots$$

The distribution of those means is the **sampling distribution**

. . .

…and it is (approximately) a *normal distribution!*

::: subpoint
○ Unimodal, symmetrical
:::

::: subpoint
○ Mean = median = mode
:::

::: subpoint
○ 68% / 95% / 99.7% within 1 / 2 / 3 SDs
:::
::::::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
x <- seq(-3.5, 3.5, length.out = 400)
plot(x, dnorm(x), type = "l", lwd = 3, col = "#377EB8", axes = FALSE,
     xlab = "", ylab = "", main = "Distribution of sample means")
axis(1, at = -3:3, labels = c(expression(mu-3*sigma), expression(mu-2*sigma), expression(mu-sigma),
                              expression(mu), expression(mu+sigma), expression(mu+2*sigma), expression(mu+3*sigma)))
for (k in 1:3) segments(c(-k, k), 0, c(-k, k), dnorm(k), lty = 2, col = "grey50")
text(0, 0.2, "68%"); text(0, 0.06, "95% within ±2σ", cex = 0.8)
```
:::
::::::::

## Picturing the Sampling Distribution {style="font-size:85%"}

::::::::: columns
::: {.column width="55%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5.5
set.seed(2)
k <- 50; n <- 20
m <- replicate(k, mean(rnorm(n, 15, 8)))
se <- 8 / sqrt(n)
lo <- m - 1.96 * se; hi <- m + 1.96 * se
miss <- lo > 15 | hi < 15
plot(m, 1:k, xlim = c(5, 25), pch = 19, col = ifelse(miss, "red", "black"),
     xlab = "Sample mean", ylab = "Sample number", main = "95% CIs from 50 samples")
segments(lo, 1:k, hi, 1:k, col = ifelse(miss, "red", "black"))
abline(v = 15, lty = 2, col = "#377EB8", lwd = 2)
text(15, k + 1.5, "true mean", col = "#377EB8", xpd = TRUE)
```
:::

::::::: {.column width="45%"}
Each dot is the mean of one sample; each line is that sample's 95% CI

Red intervals don't contain the "true" value of the mean

::::: fragment
If the dots dropped to the ground, they would pile up into a **normal distribution**

::: subpoint
○ More dots around the middle
:::

::: subpoint
○ Fewer dots toward the tails
:::
:::::

::: {style="font-size:60%"}
Simulated to mirror textbook Figure 2.8 (p. 46)
:::
:::::::
:::::::::

## What the Sampling Distribution Tells Us

It allows us to…

::: subpoint
○ specify a range of scores the population mean is expected to fall within
:::

::: subpoint
○ determine how accurate our sample mean is in estimating the true population mean
:::

::: subpoint
○ determine how likely an observed statistic (e.g., a sample mean) is, assuming some hypothesized parameter holds in the population
:::

:::: fragment
::: callout-note
If people in GA have an average IQ of 100, what is the probability that 100 randomly selected people in GA have a mean IQ of 120?
:::
::::

# Central Limit Theorem {.center}

## Central Limit Theorem (CLT)

**Definition:** if you take sufficiently large random samples from the population, then the distribution of the sample means will be approximately normally distributed

::: {style="font-size:70%"}
ⓘ It doesn't have to be a mean for the CLT to hold. It can also be a sum (a sum is part of calculating the mean!)
:::

. . .

Two things I want to point out:

1.  Repeatedly sampling from a **non-normal** population still gives you a normal sampling distribution
2.  The variance of the sampling distribution is **smaller** than the original variance

## 1) Sampling Distribution → Normal

Regardless of the population's skewness / kurtosis, the distribution of the means becomes normal

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4
set.seed(6)
pops <- list(`U-shape` = Ushape, Uniform = runif(2000, 1, 10),
             Exponential = rexp(2000, 3), `Right-skewed` = rgamma(2000, 2, 1))
par(mfrow = c(2, 4), mar = c(3, 2, 2, 1))
for (nm in names(pops)) hist(pops[[nm]], main = paste("Population:", nm), col = "#b22222",
                             border = "white", xlab = "", ylab = "", yaxt = "n")
for (nm in names(pops)) hist(replicate(2000, mean(sample(pops[[nm]], 30, replace = TRUE))),
                             main = "Sample means (n = 30)", col = "#377EB8", border = "white",
                             xlab = "", ylab = "", yaxt = "n")
```

::: subpoint
○ U-shape, uniform, exponential, extremely skewed… whatever the original distribution is
:::

## Example: U-shaped Distribution {style="font-size:85%"}

::::: columns
::: {.column width="45%"}
```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4
hist(Ushape, breaks = seq(0.75, 5.25, 0.5), col = "#4472C4", border = "white",
     main = "Population", xlab = "Value")
```

Population: $N$ = `r length(Ushape)`, $\mu$ = `r mean(Ushape)`, $\sigma$ = `r round(sd(Ushape), 2)`
:::

::: {.column width="55%"}
```{r}
#| echo: false
set.seed(12345)
tab <- t(sapply(1:10, function(i) {
  s <- sample(Ushape, 100, replace = TRUE); c(Mean = mean(s), SD = sd(s))
}))
knitr::kable(data.frame(Sample = paste("Sample", 1:10), round(tab, 2)))
```

Sampling with $n$ = 100: the 10 means average `r round(mean(tab[, "Mean"]), 3)` with an SD of only `r round(sd(tab[, "Mean"]), 3)`
:::
:::::

## Sample Size Matters

How does the sampling distribution change as $n$ (observations per sample) increases?

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 3.8
set.seed(1)
par(mfrow = c(1, 4), mar = c(4, 3, 3, 1))
hist(Ushape, breaks = seq(0.75, 5.25, 0.5), col = "#4472C4", border = "white",
     main = "Population (U-shape)", xlab = "", xlim = c(1, 5))
for (n in c(5, 30, 50)) hist(replicate(100, mean(sample(Ushape, n, replace = TRUE))),
                             xlim = c(1, 5), col = "grey80", border = "white",
                             main = paste("Sample means, n =", n), xlab = "")
```

::: fragment
The bigger $n$ is, the **narrower** the sampling distribution
:::

## Sample Size Matters: Standard Error

::::: columns
::: {.column width="60%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4.5
n <- 1:100
plot(n, sd(Ushape) / sqrt(n), xlab = "n (sample size)", ylab = "Standard error", pch = 1,
     main = "SE of the mean for the U-shaped population")
abline(v = 30, col = "red", lty = 2, lwd = 2)
```
:::

::: {.column width="40%"}
Large reductions in the standard error as $n$ increases from 1 to about 30

Beyond about $n$ = 30, there is not as much benefit to adding more subjects in terms of standard error
:::
:::::

## 2) Sampling Variance \< Original Variance

The variance of the sampling distribution of the mean is **always** smaller than the original population variance

::: subpoint
○ Less variability in the sample means than in the population
:::

::: subpoint
○ As the sample size increases, the sampling distribution becomes narrower
:::

. . .

Because the original variance is divided by $n$, the sample size:

| mean | variance | standard deviation |
|:--:|:--:|:--:|
| $\mu_{\bar{X}} = \mu$ | $\sigma^2_{\bar{X}} = \dfrac{\sigma^2}{n}$ | $\sigma_{\bar{X}} = \dfrac{\sigma}{\sqrt{n}}$ |

::: {style="font-size:80%"}
The SD of the sampling distribution is called the **standard error (SE)**

Intuition: when the sample size equals the population size, the sampling variance is essentially 0 — no error in the estimate
:::

# Sampling Distributions in R {.center}

## R Demonstrations

Two factors we're going to test:

**Different original distributions**

::: subpoint
○ If the data are not normal (e.g., uniform or exponential), is the sampling distribution still normal?
:::

**Different sample sizes**

::: subpoint
○ When each sample has 5, 30, or 200 observations
:::

## Warm-up: Functions from Earlier Labs

```{webr}
# Generate a sample
data <- c(7, 3, 5, 1, 9, 8)
sort(data)      # sort the data
length(data)    # number of data points
mean(data)
median(data)
```

## Example: U-shaped Distribution {style="font-size:85%"}

::::::: columns
::: {.column width="50%"}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 3.5
Ushape <- c(rep(5, 250), rep(4.5, 200),
            rep(4, 100), rep(3.5, 50),
            rep(3, 25), rep(2.5, 50),
            rep(2, 100), rep(1.5, 200),
            rep(1, 250))
hist(Ushape)
mean(Ushape)
sd(Ushape)
```
:::

::::: {.column width="50%"}
`rep(x, times)` repeats a value

::: subpoint
○ `rep(5, 250)` = the number 5, 250 times
:::

Type `?rep` in the console for more info

::: fragment
What does this population look like? Is it normal?
:::
:::::
:::::::

## R Functions: `sample()` {style="font-size:85%"}

Extract a sample from a given dataset

`sample(dataname, size = n, replace = T)`

::: subpoint
○ `size` = number of observations *n*
:::

::: subpoint
○ `replace = TRUE` (or `T`): sampling *with* replacement
:::

::: subpoint
○ `replace = FALSE` (or `F`): sampling *without* replacement
:::

`set.seed()` makes random results reproducible: same seed, same samples

```{webr}
set.seed(12345)
samp1 <- sample(Ushape, size = 100, replace = T)
samp2 <- sample(Ushape, 100, replace = T)   # don't need to type "size ="
mean(samp1)
mean(samp2)
```

## Building a Sampling Distribution by Hand {style="font-size:85%"}

:::::: columns
::: {.column width="55%"}
```{webr}
set.seed(12345)
samp.dist <- c(mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)),
               mean(sample(Ushape, 100, replace = T)))
hist(samp.dist, xlim = c(1, 5))
mean(samp.dist)
sd(samp.dist)   # this is the standard error (SE)!
```
:::

:::: {.column width="45%"}
10 samples of $n$ = 100, one mean each

Compare with `hist(Ushape, xlim = c(1, 5))`

::: fragment
Typing this out gets old fast…

What if we want 1,000 samples?
:::
::::
::::::

## R Functions: `for` Loops {style="font-size:85%"}

Repeat the same code many times

```{r}
#| eval: false
samples <- rep(NA, N)             # 1. "blank" vector of N placeholders
for (i in 1:N) {                  # 2. do this N times...
  samples[i] <- mean(sample(dataname, size = n, replace = T))   # 3. ...fill slot i
}
```

::: subpoint
○ **N** = number of sample sets; **n** = number of observations per sample (sample size)
:::

::: subpoint
○ `NA` is a placeholder we fill in later
:::

```{webr}
#| fig-width: 5
#| fig-height: 3
samples <- rep(NA, 10)
for (i in 1:10) {
  samples[i] <- mean(sample(Ushape, 100, replace = T))
}
samples
hist(samples, xlim = c(1, 5))
```

::: {style="font-size:70%"}
ⓘ Every run gives different values: it's random! Add `set.seed()` first to make it reproducible
:::

## Different Sample Sizes (*n*) {style="font-size:85%"}

:::::::: columns
::: {.column width="55%"}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
n <- 5   # try 10, 30, 50, 100, 200, 500!

samples_n <- rep(NA, 100)
for (i in 1:100) {
  samples_n[i] <- mean(sample(Ushape, n, replace = T))
}

hist(samples_n, xlim = c(1, 5),
     main = paste("Sample means, n =", n))
sd(samples_n)
```
:::

:::::: {.column width="45%"}
$N$ = 100 sample sets each time; only $n$ changes

Change `n` and re-run: what happens to

::: subpoint
○ the shape?
:::

::: subpoint
○ the spread (`sd()`, i.e., the SE)?
:::

::: fragment
Check: $\sigma / \sqrt{n}$ = `sd(Ushape) / sqrt(n)`
:::
::::::
::::::::

## FYI: The Number of Samplings (*N*)

The number of sample sets does **not** change the essential properties of the sampling distribution

::: subpoint
○ The sampling distribution is (hypothetically) fixed once the sample size *n* is set
:::

::: subpoint
○ More samplings just make the visualization easier
:::

```{webr}
#| fig-width: 10
#| fig-height: 3
par(mfrow = c(1, 3))
for (N in c(10, 100, 10000)) {
  m <- rep(NA, N)
  for (i in 1:N) m[i] <- mean(sample(Ushape, 100, replace = T))
  hist(m, xlim = c(2.4, 3.6), main = paste(N, "samplings"))
}
```

## R Functions: Making Distributions {style="font-size:85%"}

| Function | Generates *n* random numbers from a… |
|----|----|
| `rnorm(n, mean = , sd = )` | normal distribution with the specified mean and SD |
| `runif(n, min = , max = )` | uniform distribution with the specified min and max |
| `rexp(n, rate = )` | exponential distribution with the specified rate (λ) |

We saw `rnorm()` in Lab 4; `runif()` and `rexp()` work the same way

## Normal Distribution Sampling

:::::: columns
::: {.column width="55%"}
```{webr}
#| autorun: true
#| fig-width: 6
#| fig-height: 3
norm.dist <- rnorm(500, mean = 100, sd = 10)

sample.norm <- rep(NA, 500)
for (i in 1:500) {
  sample.norm[i] <- mean(sample(norm.dist, 30, replace = T))
}

par(mfrow = c(1, 2))
hist(norm.dist, main = "Population")
hist(sample.norm, main = "Sample means (n = 30)")
mean(sample.norm)
sd(sample.norm)
```
:::

:::: {.column width="45%"}
Population: normal, *σ* = 10

Sampling distribution: centered at the same mean, but much narrower

::: fragment
Predicted SE: $10 / \sqrt{30} \approx$ `r round(10 / sqrt(30), 2)`
:::
::::
::::::

## Uniform Distribution Sampling

::::: columns
::: {.column width="55%"}
```{webr}
#| autorun: true
#| fig-width: 6
#| fig-height: 3
uni.dist <- runif(600, min = 1, max = 10)

sample_means <- rep(NA, 1000)
for (i in 1:1000) {
  sample_means[i] <- mean(sample(uni.dist, 30, replace = T))
}

par(mfrow = c(1, 2))
hist(uni.dist, main = "Population")
hist(sample_means, main = "Sample means (n = 30)")
mean(sample_means)
sd(sample_means)
```
:::

::: {.column width="45%"}
Population: flat, not normal at all

Sampling distribution: normal!
:::
:::::

## Exponential Distribution Sampling

:::::: columns
::: {.column width="55%"}
```{webr}
#| autorun: true
#| fig-width: 6
#| fig-height: 3
exp.dist <- rexp(1000, rate = 3)

sample.exp <- rep(NA, 1000)
for (i in 1:1000) {
  sample.exp[i] <- mean(sample(exp.dist, 40, replace = T))
}

par(mfrow = c(1, 2))
hist(exp.dist, main = "Population")
hist(sample.exp, main = "Sample means (n = 40)")
mean(sample.exp)
sd(sample.exp)
```
:::

:::: {.column width="45%"}
Population: extremely right-skewed

Sampling distribution: still normal!

::: fragment
Try `n = 5` instead of 40: is it still normal? (Remember "sufficiently large")
:::
::::
::::::

## Must-Take-Home Message {style="font-size:90%"}

**Population distribution**

::: subpoint
○ The distribution of the population of research interest
:::

::: subpoint
○ Unknown, so estimated with statistics, but assumed to be fixed
:::

**Sample distribution**

::: subpoint
○ The distribution of the dataset we get from experiments/surveys
:::

::: subpoint
○ Used to infer the population parameters
:::

**Sampling distribution**

::: subpoint
○ Distribution of the MEANS of samples
:::

::: subpoint
○ CLT: approximately NORMALLY distributed, if certain conditions are met (e.g., *n* \> 30)
:::

::: subpoint
○ Its SD is the standard error: $\sigma / \sqrt{n}$
:::

## Lab 6 Assignment!

<!-- Add assignment reminders here. -->

-   Look back at the slides and the live code to help you work through this week's assignment
-   Read the instructions carefully, and check your work before you turn it in!
-   Please submit both the knitted file and the .Rmd on Canvas!
