---
title: "Null Hypothesis Significance Testing I"
subtitle: "PSYC 2020-A01 / PSYC 6022-A01 | Lab 7"
author: "Shuheng Jiang"
format:
  live-revealjs:
    slide-number: true
    show-slide-number: all
    css: styles.css
    width: 1200
engine: knitr
execute:
  echo: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{webr}
#| edit: false
#| output: false
# hidden setup: a helper that draws the standard normal curve and shades the rejection region(s)
draw_z <- function(z = NULL, tail = "two", alpha = .05, main = "") {
  x <- seq(-4, 4, length.out = 400)
  plot(x, dnorm(x), type = "l", lwd = 2, xlab = "z", ylab = "", yaxt = "n", main = main)
  shade <- function(a, b) polygon(c(a, seq(a, b, length.out = 100), b),
                                  c(0, dnorm(seq(a, b, length.out = 100)), 0),
                                  col = adjustcolor("firebrick", .4), border = NA)
  if (tail == "two")   { cv <- qnorm(1 - alpha / 2); shade(-4, -cv); shade(cv, 4) }
  if (tail == "lower") { cv <- qnorm(alpha); shade(-4, cv) }
  if (tail == "upper") { cv <- qnorm(1 - alpha); shade(cv, 4) }
  if (!is.null(z)) abline(v = z, col = "#377EB8", lwd = 3)
}
```

```{r}
#| include: false
draw_z <- function(z = NULL, tail = "two", alpha = .05, main = "") {
  x <- seq(-4, 4, length.out = 400)
  plot(x, dnorm(x), type = "l", lwd = 2, xlab = "z", ylab = "", yaxt = "n", main = main)
  shade <- function(a, b) polygon(c(a, seq(a, b, length.out = 100), b),
                                  c(0, dnorm(seq(a, b, length.out = 100)), 0),
                                  col = adjustcolor("firebrick", .4), border = NA)
  if (tail == "two")   { cv <- qnorm(1 - alpha / 2); shade(-4, -cv); shade(cv, 4) }
  if (tail == "lower") { cv <- qnorm(alpha); shade(-4, cv) }
  if (tail == "upper") { cv <- qnorm(1 - alpha); shade(cv, 4) }
  if (!is.null(z)) abline(v = z, col = "#377EB8", lwd = 3)
}
```

## Quiz 5 Review


::: callout-note
Random samples of $n$ = 9 scores are selected from a population that is **strongly positively skewed**. What can be said about the shape of the distribution of sample means?
:::

::: fragment
**The shape cannot be assumed to be normal.** It will most likely still be positively skewed (though less skewed than the population)
:::

::::: fragment
Why? The Central Limit Theorem needs either

::: subpoint
○ a **normal** population (then any $n$ works), or
:::

::: subpoint
○ a **large** sample, roughly $n \ge 30$
:::

Here we have neither: a skewed population **and** a small $n$ = 9
:::::

## Quiz 5 Review: Let's Check in R

:::::: columns
::: {.column width="60%"}
```{webr}
#| autorun: true
#| fig-width: 6
#| fig-height: 3.5
set.seed(5)
pop <- rexp(10000, rate = 1)   # strongly positively skewed
n <- 9                          # try 30, then 100!

means <- rep(NA, 2000)
for (i in 1:2000) {
  means[i] <- mean(sample(pop, n, replace = T))
}

par(mfrow = c(1, 2))
hist(pop, main = "Population", xlab = "")
hist(means, main = paste("Sample means, n =", n), xlab = "")
```
:::

:::: {.column width="40%"}
Same `for` loop as Lab 6

With $n$ = 9, the sample means still have a long right tail

::: fragment
Change `n` to 30 and 100 and re-run: the tail shrinks and the shape becomes normal. That is the CLT at work
:::
::::
::::::

## Outline

-   From Sampling Distributions to Hypothesis Testing
-   The *z*-Distribution in R
-   Hypothesis Testing Step by Step
-   One-Tailed vs. Two-Tailed Tests

Learning objectives: **R:** `pnorm()`, `qnorm()`, `lower.tail`, writing a *z*-test by hand

**Statistics:** *z*-test for a sample mean, *p*-values, critical values, reject vs. retain

# From Sampling Distributions to Hypothesis Testing {.center}

## Quick Review: Sampling Distributions

From Lab 6: the sampling distribution of the mean has

|         mean          |                standard error                 |
|:---------------------:|:---------------------------------------------:|
| $\mu_{\bar{X}} = \mu$ | $\sigma_{\bar{X}} = \dfrac{\sigma}{\sqrt{n}}$ |

. . .

Learning check: samples of $n$ = 36 from a population with $\mu$ = 80, $\sigma$ = 12

```{webr}
mu <- 80
sigma <- 12
n <- 36

sigma / sqrt(n)   # standard error
```

## Two Distributions, Two *z*-Scores

::::::::::: columns
:::::: column
**Distribution of the data**

::: subpoint
○ Score: $x$
:::

::: subpoint
○ Mean: $\mu$, SD: $\sigma$
:::

::: subpoint
○ $z = \dfrac{x - \mu}{\sigma}$ (Lab 4!)
:::
::::::

:::::: column
**Sampling distribution**

::: subpoint
○ Score: $\bar{x}$
:::

::: subpoint
○ Mean: $\mu$, SE: $\sigma / \sqrt{n}$
:::

::: subpoint
○ $z_{\bar{x}} = \dfrac{\bar{x} - \mu}{\sigma / \sqrt{n}}$
:::
::::::
:::::::::::

. . .

The second one is our **test statistic**: how many standard errors is our sample mean away from the hypothesized population mean?

## Seeing the Test Statistic {style="font-size:85%"}

Population under $H_0$: $\mu$ = 10, $\sigma$ = 2. We collect a sample of $n$ = 20 with $\bar{x}$ = 12

::::::::: columns
::: {.column width="60%"}
```{webr}
#| autorun: true
#| fig-width: 6
#| fig-height: 3.5
pop_mean <- 10
pop_sd   <- 2
n_samp   <- 20
samp_mean <- 12

z <- (samp_mean - pop_mean) / (pop_sd / sqrt(n_samp))
z
draw_z(z, main = paste("z =", round(z, 2)))
```
:::

::::::: {.column width="40%"}
The curve is the sampling distribution **if** $H_0$ were true; the blue line is where our sample lands

Change one number at a time and re-run:

::: subpoint
○ `pop_sd <- 8` (noisier population)
:::

::: subpoint
○ `n_samp <- 200` (bigger sample)
:::

::: subpoint
○ `samp_mean <- 20`, `n_samp <- 5` (big difference, tiny sample)
:::

::: fragment
What makes $z$ bigger? Smaller?
:::
:::::::
:::::::::

# The *z*-Distribution in R {.center}

## The `norm()` Family

A standard normal distribution *is* the *z*-distribution, so the `norm()` functions (Lab 4) do all the work

| Function   | Gives you…                                  |
|------------|---------------------------------------------|
| `rnorm(n)` | *n* random *z*-values                       |
| `pnorm(z)` | the probability **at or below** a *z*-value |
| `qnorm(p)` | the *z*-value **at** a given probability    |

::: subpoint
○ Default is `mean = 0, sd = 1`, so we don't need to type them
:::

## `pnorm()`: From *z* to Probability

```{webr}
pnorm(-1)                     # area below z = -1
pnorm(1.23, lower.tail = F)   # area above z = 1.23
pnorm(c(-1.96, 1.96))
```

. . .

::: callout-note
`pnorm()` gives the area **at or below** the value. For a positive *z*, the *p*-value is the area **above** it, so use `lower.tail = FALSE`
:::

These match the lecture: *z* = −1.00 → .16; *z* = 1.23 → .11

## `qnorm()`: From Probability to *z* (Critical Values)

Where does ±1.96 come from?

```{webr}
qnorm(c(.025, .975))   # two-tailed, alpha = .05: 2.5% in each tail
qnorm(.05)             # one-tailed (lower), alpha = .05
qnorm(.95)             # one-tailed (upper), alpha = .05
```

. . .

::: subpoint
○ Two-tailed: $z_{crit}$ = ±1.96
:::

::: subpoint
○ One-tailed: $z_{crit}$ = −1.65 (left) or +1.65 (right)
:::

## Getting a *p*-Value for Any *z*

:::::: columns
::: {.column width="55%"}
```{webr}
#| autorun: true
#| fig-width: 5.5
#| fig-height: 3.5
z <- -1.23

# one-tailed: just the tail z falls in
p_one <- pnorm(-abs(z))
# two-tailed: both tails
p_two <- 2 * pnorm(-abs(z))

c(one_tailed = p_one, two_tailed = p_two)
draw_z(z, tail = "two")
```
:::

:::: {.column width="45%"}
`-abs(z)` always puts us in the lower tail, so this works for positive **and** negative *z*

Two-tailed *p* = 2 × one tail

::: fragment
Try `z <- 3`, `z <- 0`, `z <- 1.96`

What *p*-value do you get at exactly 1.96?
:::
::::
::::::

# Hypothesis Testing Step by Step {.center}

## The Logic of Hypothesis Testing

**1. State the hypotheses**

::: subpoint
○ $H_0$: no effect / no difference (e.g., $\mu$ = 100)
:::

::: subpoint
○ $H_1$: there is an effect / a difference (e.g., $\mu \neq$ 100)
:::

**2. Set the significance level**: $\alpha$ = .05, and find the critical value

**3. Compute the test statistic** ($z$) and its *p*-value, *assuming* $H_0$ is true

**4. Make a decision**

. . .

|   | Decision |
|------------------------------------|------------------------------------|
| $p \le \alpha$, or $|z| \ge |z_{crit}|$ | **Reject** $H_0$: statistically significant |
| $p > \alpha$, or $|z| < |z_{crit}|$ | **Fail to reject** (retain) $H_0$: not significant |

::: {style="font-size:70%"}
ⓘ We never say we "accept" $H_0$ or that $H_0$ is true
:::

## Example: Weight of Cats {style="font-size:85%"}

Normal cats: $\mu$ = 100 g, $\sigma$ = 40. A sample of $n$ = 16 cats with prenatal alcohol exposure has $\bar{x}$ = 70. Does alcohol affect birth weight?

::::::: columns
::: {.column width="55%"}
```{webr}
#| autorun: true
#| fig-width: 5.5
#| fig-height: 3
# Step 1: H0: mu = 100   H1: mu != 100
# Step 2
alpha  <- .05
z_crit <- qnorm(1 - alpha / 2)
# Step 3
z <- (70 - 100) / (40 / sqrt(16))
p <- 2 * pnorm(-abs(z))
c(z = z, z_crit = z_crit, p = p)
# Step 4
p <= alpha
draw_z(z, tail = "two")
```
:::

::::: {.column width="45%"}
::: fragment
$z$ = −3 is far in the rejection region

$p$ \< .05 → **reject** $H_0$

Prenatal alcohol exposure has a statistically significant effect on cats' birth weight
:::

::: {style="font-size:65%"}
ⓘ The lecture reports the one-tail area, `pnorm(-3)` = .00135; the two-tailed *p* doubles it (.0027). Same decision either way
:::
:::::
:::::::

## Your Turn: The "Evil" Professor {style="font-size:85%"}

Your professor believes extra homework **changes** final exam scores. Population: $\mu$ = 70, $\sigma$ = 15. A sample of $n$ = 25 students who got 50 extra HW questions a week scored $\bar{x}$ = 75.

```{webr}
# a. H0:            H1:
# b. critical value at alpha = .05 (two-tailed)

# c. test statistic

# d. decision and conclusion

```

::: {.fragment style="font-size:80%"}
Check: $z = \dfrac{75 - 70}{15 / \sqrt{25}} = \dfrac{5}{3}$ = 1.67, which is less than 1.96 → **retain** $H_0$
:::

## Example *z*-Test with Real Data {style="font-size:80%"}

A flower researcher tells us the population mean of iris petal **length** is 4 cm and of petal **width** is 3 cm, both with $\sigma^2$ = 4. Do the `iris` sample means differ?

::::: columns
::: column
```{webr}
samp_mean <- mean(iris$Petal.Length)
pop_mean  <- 4
pop_sd    <- sqrt(4)
samp_n    <- sum(!is.na(iris$Petal.Length))

z_length <- (samp_mean - pop_mean) /
  (pop_sd / sqrt(samp_n))
z_length
2 * pnorm(-abs(z_length))
```
:::

::: column
```{webr}
# your turn: petal width (pop mean = 3)

```

`sum(!is.na(x))` counts only the non-missing values, which is safer than `length()` or `nrow()` if there are `NA`s
:::
:::::

::: fragment
Interpretation? Which test is significant?
:::

# One-Tailed vs. Two-Tailed Tests {.center}

## Directional vs. Non-Directional Hypotheses {style="font-size:85%"}

::::::::::: columns
:::::: column
**Non-directional** → two-tailed test

"There is a **difference**…"

::: subpoint
○ $H_1$: $\mu \neq \mu_0$
:::

::: subpoint
○ $\alpha$/2 = .025 in **each** tail
:::

::: subpoint
○ $z_{crit}$ = ±1.96
:::
::::::

:::::: column
**Directional** → one-tailed test

"…**increases** / **decreases** / **better than**…"

::: subpoint
○ $H_1$: $\mu > \mu_0$ (upper) or $\mu < \mu_0$ (lower)
:::

::: subpoint
○ all of $\alpha$ = .05 in **one** tail
:::

::: subpoint
○ $z_{crit}$ = +1.65 or −1.65
:::
::::::
:::::::::::

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 2.6
par(mfrow = c(1, 3), mar = c(4, 1, 2, 1))
draw_z(tail = "two",   main = "Two-tailed")
draw_z(tail = "lower", main = "Lower-tailed")
draw_z(tail = "upper", main = "Upper-tailed")
```

## Practice: Directional or Non-Directional?

1.  There is a difference in body weights, with men weighing more than women.
2.  Women are better at concentrating on more than one thing at a time than men.
3.  There is a relationship between studying and final exam score.
4.  As employees earn more money, they become less likely to leave for a new job.

::: fragment
Directional: 1, 2, 4 \| Non-directional: 3
:::

## The "Evil" Professor, Revisited {style="font-size:85%"}

Now your professor believes extra homework **increases** final exam scores (same data: $\mu$ = 70, $\sigma$ = 15, $n$ = 25, $\bar{x}$ = 75)

::::::::: columns
::: {.column width="55%"}
```{webr}
#| autorun: true
#| fig-width: 5.5
#| fig-height: 3
z <- (75 - 70) / (15 / sqrt(25))

# two-tailed (H1: mu != 70)
c(z_crit = qnorm(.975), p = 2 * pnorm(-abs(z)))

# upper-tailed (H1: mu > 70)
c(z_crit = qnorm(.95), p = pnorm(z, lower.tail = F))

draw_z(z, tail = "upper")
```
:::

::::::: {.column width="45%"}
Same data, same $z$ = 1.67…

::::: fragment
::: subpoint
○ Two-tailed: 1.67 \< 1.96 → retain $H_0$
:::

::: subpoint
○ Upper-tailed: 1.67 \> 1.65 → **reject** $H_0$
:::
:::::

::: fragment
This is why the direction must be decided **before** looking at the data!
:::
:::::::
:::::::::

## Must-Take-Home Message {style="font-size:90%"}

$$z_{\bar{x}} = \frac{\bar{x} - \mu}{\sigma / \sqrt{n}}$$

::: subpoint
○ `pnorm()` turns a *z* into a probability; use `lower.tail = F` or `-abs(z)` for the upper tail
:::

::: subpoint
○ `qnorm()` turns $\alpha$ into a critical value: ±1.96 (two-tailed), ±1.65 (one-tailed)
:::

::: subpoint
○ Two-tailed *p* = 2 × one-tailed *p*
:::

::: subpoint
○ $p \le \alpha$ → reject $H_0$; $p > \alpha$ → fail to reject (never "accept") $H_0$
:::

::: subpoint
○ Directional hypothesis → one-tailed test; decide before collecting data
:::

## Lab 7 Assignment!

<!-- Add assignment reminders here. -->

-   Look back at the slides and the live code to help you work through this week's assignment
-   Please submit both the knitted file and the .Rmd on Canvas!
