---
title: "Central Tendency"
subtitle: "PSYC 2020-A01 / PSYC 6022-A01 | Lab 2"
author: "Shuheng Jiang"
format:
  live-revealjs:
    css: styles.css
engine: knitr
execute:
  echo: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

## Absence and Late Submission Policy
::: {style="font-size: 70%"}

- Homework and quizzes will generally not be reopened once they have closed and been graded, except in exceptional circumstances.

- For an unavoidable absence, students may submit an Institute Approved Absence (IAA) request through Georgia Tech. Approved IAA requests will be honored.

- Alternatively, students may provide appropriate supporting documentation, such as a doctor’s or dentist’s note. Such documentation will generally be accepted for absence or late-submission accommodations.

- For non-emergency circumstances that you know about in advance, please contact me at least 3 days before the affected lab or assignment deadline. For emergencies or unexpected situations, please contact me as soon as reasonably possible.

:::

## Homework 1 
Everyone has done a great job! 
remember to set up the second level header in the white text space, not within the gray code chunk 


## Review of Central Tendency!

Mean: [Sum of all values divided by the total number of values]{.fragment fragment-index="0"}

Median: [When sorted lowest to highest, the middle value]{.fragment fragment-index="1"}

Mode: [The value that appears most often]{.fragment fragment-index="2"}


## Central Tendency in R: Mean

We can calculate central tendencies in two ways:

Given this dataset, calculate the mean

```{r}
c(2, 3, 12, 4, 4)
```

:::: {.columns}
::: {.column}
By hand (computer)

```{webr}
# let's calculate the mean!
```
:::

::: {.column}
With the `mean(x)` function

::: {.subpoint}
○ `x` = vector of data
:::

```{webr}
# let's calculate the mean!
```
:::
::::

## Central Tendency in R: Median

Given this dataset, calculate the median

```{r}
c(2, 3, 12, 4, 4)
```

:::: {.columns}
::: {.column}
By hand (computer)

```{webr}
# let's calculate the median!
```
:::

::: {.column}
With the `median(x)` function

::: {.subpoint}
○ `x` = vector of data
:::

```{webr}
# let's calculate the median!
```
:::
::::

## Central Tendency in R: Mode

Given this dataset, calculate the mode

```{r}
c(2, 3, 12, 4, 4)
```

With the `mode()` function

```{webr}
# let's calculate the mode!
```

::: {.fragment fragment-index="0"}
Doesn't work :(

Have to create our own
:::

## R Functions

We've seen some built-in R functions (e.g., `mean()`, `median()`), but we can also make our own

:::: {.columns}
::: {.column}
```
function_name <- function(argument) {
  do some stuff
  return(this stuff)
}
```
:::

::: {.column style="font-size:70%"}
ⓘ Don't actually need to call `return()`; R will automatically return the last expression
:::
::::

::: {.fragment fragment-index="0"}
Then, you can call the function

`function_name(specific_argument)`
:::

::: {.fragment fragment-index="1"}
To keep the results, make sure to assign them to some variable

`very_important_results <- function_name(specific_argument)`
:::

## Central Tendency in R: Mode

Given this dataset, calculate the mode

```{r}
c(2, 3, 12, 4, 4)
```

```{r}
mode_func <- function(x) {
  sort(table(x), decreasing = T)[1]
}
```

How does this work?

```{webr}
# let's calculate the mode!
```

## Measures of Variability in R 

Variance: [Average of the squared distances of each value from the mean $s^2 = \dfrac{\sum (x_i - \bar{x})^2}{n - 1}$]{.fragment fragment-index="0"}

Standard deviation: [Square root of the variance; the typical distance of a value from the mean, in the original units]{.fragment fragment-index="1"}

Quantile: [A cut point that splits sorted data into equal-sized groups (e.g., 25% of values fall below the 1st quartile, 75% below the 3rd)]{.fragment fragment-index="2"}

IQR: [3rd quartile minus 1st quartile; the range of the middle 50% of the data]{.fragment fragment-index="3"}

## Variability in R: Variance, SD, IQR


Given this dataset, calculate the median

```{r}
c(2, 3, 12, 4, 4)
```

With the `var(x)` function

○ `x` = vector of data

```{webr}
# let's calculate the variance!
```

With the `sd(x)` function

```{webr}
# let's calculate the standard deviation!
```

With the `IQR(x)` function

```{webr}
# let's calculate the IQR!
```


## Descriptive Statistics in R

Takes time to look at all these for a lot of variables, even with functions

The `summary(object)` function provides us a quick overview of this information

::: {.subpoint}
○ `object` = for our purposes, a dataframe
:::

```{webr}
# let's get descriptive statistics for the iris dataset
```

What all do we get?

::: {.fragment fragment-index="1"}
Minimum and maximum

1st quartile, median, 3rd quartile

Mean
:::

## Descriptive Statistics in R: 

`summary()` gives us the basics, but no variance, SD, or skew

The `describe(object)` function from the `psych` package gives us more

::: {.subpoint}
○ `object` = a dataframe (or a single variable)
:::

Do need to install the `psych` package first

```{r}
#| echo: true
#| eval: false
install.packages("psych")
```

```{webr}
library(psych)
# let's describe the iris dataset
```

::: {.fragment fragment-index="1"}
What all do we get?

n, mean, sd, median, trimmed mean, mad, min, max, range, skew, kurtosis, se
:::

## Visualizations

Don't rush: graph your data!

What should graphs do?

::: {.subpoint}
○ Show the data
:::

::: {.subpoint}
○ Draw the reader primarily to the data (not the graphical effects)
:::

::: {.subpoint}
○ Avoid distorting the data
:::

::: {.subpoint}
○ Present many numbers with minimum ink
:::

::: {.subpoint}
○ Make large data sets coherent
:::

::: {.subpoint}
○ Encourage the reader to compare different pieces of data
:::

## Visualizations: Histograms

:::: {.columns}
::: {.column}
An example of (simulated) SAT scores

What do we see here?

::: {.fragment fragment-index="0"}
::: {.subpoint}
○ Outliers at zero! Not a possible SAT score
:::
:::

::: {.fragment fragment-index="1"}
::: {.subpoint}
○ Negatively skewed: more data on the left than on the right
:::
:::
:::

::: {.column}
```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 4
set.seed(2020)
sat <- round(1600 - rgamma(500, shape = 2, scale = 120))
sat <- pmin(pmax(sat, 400), 1600)
sat <- c(sat, rep(0, 6))
hist(sat, breaks = 20, col = "grey80", border = "white",
     main = "Simulated SAT scores", xlab = "SAT score")
```
:::
::::

## Skew

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3
#| fig-align: center
par(mfrow = c(1, 2), mar = c(3, 1, 2, 1))
x <- seq(0, 1, length.out = 400)
plot(x, dbeta(x, 2, 6), type = "l", lwd = 3, col = "#377EB8",
     axes = FALSE, xlab = "", ylab = "", main = "Positive skew")
abline(v = qbeta(0.5, 2, 6), lty = 2)
plot(x, dbeta(x, 6, 2), type = "l", lwd = 3, col = "#E41A1C",
     axes = FALSE, xlab = "", ylab = "", main = "Negative skew")
abline(v = qbeta(0.5, 6, 2), lty = 2)
```

:::: {.columns}
::: {.column}
Positive skew, right-tailed

The mass of the distribution is concentrated on the left of the figure
:::

::: {.column}
Negative skew, left-tailed

The mass of the distribution is concentrated on the right of the figure
:::
::::

## Skewness Demonstration

:::: {.columns}
::: {.column width="80%"}
<iframe width="90%" height="600px" src="https://fabiosetti.shinyapps.io/Skew_Normal_distribution/"></iframe>
:::

::: {.column width="20%"}
Full screen version [here](https://fabiosetti.shinyapps.io/Skew_Normal_distribution/){target="_blank"}

<br>

Skewness demonstration!

Credits to Fabio Setti
:::
::::

## Let's Do Some Visualization

### Base R Graphics

:::: {.columns}
::: {.column .fragment fragment-index="0"}
```{webr}
#| fig-width: 5
#| fig-height: 4
plot(iris$Sepal.Length, iris$Sepal.Width)
```
:::

::: {.column .fragment fragment-index="1"}
What do we think about this?
:::
::::

## Let's Do Some Visualization

### Base R Graphics: Histogram

:::: {.columns}
::: {.column width="60%"}
`hist()` function

::: {.subpoint}
**Required arguments:**
:::

::: {.subsubpoint}
○ `x` = vector (variable) you want to plot (remember the $ function!)
:::

::: {.subpoint}
**Optional arguments:**
:::

:::: {.columns}
::: {.column}
::: {.subsubpoint}
○ `breaks` = bin count
:::

::: {.subsubpoint}
○ `main` = title
:::

::: {.subsubpoint}
○ `xlab` = label for x-axis
:::

::: {.subsubpoint}
○ `ylab` = label for y-axis
:::
:::

::: {.column}
::: {.subsubpoint}
○ `col` = color for bars
:::

::: {.subsubpoint}
○ `xlim` = range for x-axis
:::

::: {.subsubpoint}
○ `ylim` = range for y-axis
:::

::: {.subsubpoint}
○ `prob` = `T`/`F`, proportion instead of frequency
:::
:::
::::

::: {style="font-size:70%"}
If you do not set specific values for non-essential subarguments, it will use the default
:::
:::

::: {.column width="40%"}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
hist(iris$Sepal.Length,
     breaks = 7,
     main = "Histogram of Sepal Length",
     xlab = "Sepal Length",
     xlim = c(3, 9),
     border = "white",
     col = "skyblue")
```
:::
::::

## Let's Do Some Visualization

### Base R Graphics: Bar Chart

:::: {.columns}
::: {.column width="60%"}
Useful to show frequency/percentage of **categorical** data

::: {.subpoint}
○ Categorical data: not numerically meaningful, it's a category of a type
:::

X-axis displays the variable; Y-axis displays frequency

Bars **do not touch** and have equal widths

Should not be 3-dimensional!

`barplot()` function

::: {.subsubpoint}
○ `height` = a table of counts (use `table()` first!)
:::

::: {.subsubpoint}
○ `main`, `xlab`, `ylab`, `col` work the same as `hist()`
:::
:::

::: {.column width="40%"}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
barplot(table(iris$Species),
        main = "How many of each species?",
        xlab = "Species",
        ylab = "Frequency",
        col = "skyblue")
```
:::
::::

## Histogram vs. Bar Chart

:::: {.columns}
::: {.column}
**Histogram**

::: {.subpoint}
○ Interval/ratio (numeric) data
:::

::: {.subpoint}
○ Bars touch: continuity of the data
:::

::: {.subpoint}
○ `hist(x)`
:::
:::

::: {.column}
**Bar chart**

::: {.subpoint}
○ Categorical data
:::

::: {.subpoint}
○ Bars do not touch: separate categories
:::

::: {.subpoint}
○ `barplot(table(x))`
:::
:::
::::

Try it: make a bar chart of `mtcars$cyl` (number of cylinders)

```{webr}
# let's make a bar chart!
```

## Let's Do Some Visualization

### Base R Graphics: Pie Chart

:::: {.columns}
::: {.column width="60%"}
Categorical data **only**

Never use with more than 2–3 categories

Use bar charts for comparisons

If unsure, don't use a pie chart!!

::: {.fragment fragment-index="0"}
Why?

::: {.subpoint}
○ Eye is good at judging relative *linear* measures, not relative *areas*
:::

::: {.subpoint}
○ Anything a pie chart can show, a bar chart can show better (Cleveland, 1985)
:::
:::

`pie()` function

::: {.subsubpoint}
○ `x` = a table of counts, just like `barplot()`
:::
:::

::: {.column width="40%"}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
pie(table(mtcars$am),
    labels = c("Automatic", "Manual"),
    main = "Transmission type",
    col = c("skyblue", "salmon"))
```
:::
::::

## Pie Chart vs. Bar Chart

Same data, two graphs: which one is easier to read?

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
par(mfrow = c(1, 2))
pie(table(mtcars$carb), main = "Pie: number of carburetors")
barplot(table(mtcars$carb), main = "Bar: number of carburetors",
        xlab = "Carburetors", ylab = "Frequency", col = "skyblue")
```

::: {.fragment fragment-index="0"}
Take away: **don't use a pie chart, use a bar chart** for categorical data

If you use a pie chart, only when there are two categories
:::

```{webr}
# try it yourself: pie(table(mtcars$carb)) vs. barplot(table(mtcars$carb))
```

## Let's Do Some Visualization

### Base R Graphics: Line Graph

:::: {.columns}
::: {.column width="60%"}
Usually displays a variable **over the passage of time** (x-axis)

Time series graph: shows how things **change**

::: {.subpoint}
○ Long term trends
:::

::: {.subpoint}
○ Seasonal components: same pattern every year
:::

::: {.subpoint}
○ Irregular cycles and random fluctuations
:::

`plot()` function, with one new argument

::: {.subsubpoint}
○ `type = "l"` = line (`"b"` = both points and lines)
:::
:::

::: {.column width="40%"}
```{webr}
#| autorun: true
#| fig-width: 5
#| fig-height: 4
year <- 1978:2003
cases <- c(8, 6, 4, 3, 2, 2.5, 2, 1.5, 1.5, 5.5, 2, 2.2, 2.1, 1.8,
           1.5, 1, 0.8, 0.6, 0.5, 0.4, 0.3, 0.3, 0.2, 0.2, 0.1, 0.1)
plot(year, cases, type = "b",
     main = "Mumps by year (per 100,000)",
     xlab = "Year", ylab = "Incidence")
```
:::
::::

## Line Graph Practice

What do we see in this graph?

::: {.fragment fragment-index="0"}
::: {.subpoint}
○ Overall decreasing trend (the vaccine!)
:::
:::

::: {.fragment fragment-index="1"}
::: {.subpoint}
○ Small peak in 1987: irregular pattern or random fluctuation
:::
:::

Try it: R has a built-in monthly time series called `AirPassengers`

```{webr}
# let's make a line graph!
plot(AirPassengers, type = "l")
```

::: {.fragment fragment-index="2"}
Trend? Seasonal component?
:::

## Let's Do Some Visualization

### Base R Graphics: Stem-and-Leaf Plot

:::: {.columns}
::: {.column width="60%"}
Visual display that **orders** numbers while also displaying **shape**

**All** data points are included

To create:

::: {.subpoint}
○ Order data
:::

::: {.subpoint}
○ Create stems (the leading digit)
:::

::: {.subpoint}
○ Attach leaves (the last digit)
:::

`stem()` function

::: {.subsubpoint}
○ `x` = vector of data
:::

::: {.subsubpoint}
○ `scale` = how many stems (2 = split each stem in two)
:::
:::

::: {.column width="40%"}
```{webr}
#| autorun: true
scores <- c(44, 46, 47, 49, 63, 64, 66, 68, 68,
            72, 72, 75, 76, 81, 84, 88, 106)
stem(scores)
```
:::
::::

## Stem-and-Leaf Practice

Read the plot: what is the smallest score? The most common stem?

```{webr}
scores <- c(44, 46, 47, 49, 63, 64, 66, 68, 68,
            72, 72, 75, 76, 81, 84, 88, 106)
# let's make a stem-and-leaf plot!
```

::: {.fragment fragment-index="0"}
Now try `stem(scores, scale = 2)`: what changed?
:::

::: {.fragment fragment-index="1"}
Bonus: compare `stem(scores)` with `hist(scores)`. What does the stemplot show that the histogram hides?
:::

## Lab Assignmnt 2
Download the dataset
save them in the same folder as your rmd. file 
set the folder directory as working directory 
```{r}
#| eval: false
#| echo: true
dataframe <- read.csv("file_name")
```

